---
title: "Research on NFL data"
author: "Team 2 (Patrick Berthelsen, Natascha Svarre, Alex Rye & Nino Krupic)"
date: "21/11/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Introduction
## 2 Theory
## 3 Data 

```{r}

pacman::p_load(
  tidyverse, #tidyverse includes ggplot2, dplyr, forcats and more. 
  caret, 
  ggvis,
  rpart.plot,
  ggmap, #for plotting maps
  RColorBrewer, #provide pretty colour palettes
  lubridate, #for converting time variables into date format
  hrbrthemes, #for more style themes
  data.table, #for working with data.table
  tidytext, #for text mining functions such as unnest()
  wordcloud,
  tm, #for removing stopwords
  ggdist, #for raincloud plots
  tidyquant, #for theme_tq
  prismatic, #for color handling in plots
  gganimate, #for animated plots
  gifski, #for rendering animated plots as GIFs instead of single frames
  knitr,
  naivebayes,
  randomForest,
  tree,
  devtools,
  plotrix,
  reprtree, 
  dynlm
)


```


### 3.1 Data cleaning

#### 3.1.1 Data on NFL season standings 2000-2019
```{r}
standings <- read.csv("standings.csv")
standings <- as.data.frame(standings)

sb_winners <- standings[ which(standings$sb_winner=="Won Superbowl"), ] #Subset of data including only SB-winners

standings[ ,c('year', 'team')] <- list(NULL) #deleting irrelavent variables

standings$playoffs <- ifelse(standings$playoffs == "Playoffs", 1 ,0)
standings$sb_winner <- ifelse(standings$sb_winner == "Won Superbowl", 1 ,0)

str(standings)
```

#### 3.1.2 Data on NFL game results 2000-2019
```{r}
games <- read.csv("games.csv")
games <- as.data.frame(games)
games$winners <- stringr::word(games$winner, -1)
games[ ,c('winner', 'home_team', 'away_team', 'day', 'date', 'time', 'home_team_city', 'away_team_city', 'year', 'week')] <- list(NULL) #deleting irrelavent variables
games$tie <- ifelse(games$tie == "NA", 0 ,1)
games$tie[is.na(games$tie)] <- 0
games$winners <- ifelse(games$tie == 1, "No Winner", games$winners)
games$home_win <- ifelse(games$winners == games$home_team_name ,1,0)

games$pts_home <- ifelse(games$home_win == 1, games$pts_win, games$pts_loss) #creating variable: home team points
games$pts_away <- ifelse(games$home_win == 0, games$pts_win, games$pts_loss) #creating variable: away team points 

games$yds_home <- ifelse(games$home_win == 1, games$yds_win, games$yds_loss) #creating variable: home team yards
games$yds_away <- ifelse(games$home_win == 0, games$yds_win, games$yds_loss) #creating variable: away team yards

games$turnovers_home <- ifelse(games$home_win == 1, games$turnovers_win, games$turnovers_loss) #creating variable: home team turnovers
games$turnovers_away <- ifelse(games$home_win == 0, games$turnovers_win, games$turnovers_loss) #creating variable: away team turnovers

games[ ,c('pts_win', 'pts_loss', 'yds_win', 'yds_loss', 'turnovers_win', 'turnovers_loss')] <- list(NULL) #deleting old variables

str(games)


```


## 4 Hypotheses

## 5 Analysis
### 5.1 EDA 
```{r}

#Barplot of wins in regular season by the SB-winner 2000-2019
sb_winners %>% 
  ggplot(aes(x = year, y = wins)) + 
  geom_line() +
  labs(title = "Regular season wins by SB-winners 2000-2019", x = "Year", y = "Games won") +
  theme_linedraw() + 
  theme(plot.title = element_text(hjust = 0.5))

#Barplot of SB-wins per team
ggplot(data = sb_winners) + 
  geom_bar(aes(team_name)) + 
  theme_linedraw() + 
  labs(title = "Number of Superbowl won pr team", x = "Teams", y = "SB won") +
  theme(plot.title = element_text(hjust = 0.5))

#Barplot of total wins per team
ggplot(data = games) + 
  geom_bar(aes(x = fct_infreq(winners))) + 
  theme_linedraw() + 
  labs(title = "Number of games won pr team", x = "Teams", y = "Games won") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  coord_flip() 



  #Correlation between points and yards
games %>% 
    ggplot(aes(x = yds_home, y = pts_home)) + 
    geom_point()

games %>% 
    ggplot(aes(x = yds_away, y = pts_away)) + 
    geom_point()


```


#### 5.1.1 Wordcloud
```{r}
#Creating dataset of only winners of games
game_winners <- games["winners"] %>% 
  group_by(winners) %>% 
  mutate(total_wins = n()) %>% 
  ungroup() %>% 
  count(winners, sort = TRUE)

wordcloud(words = game_winners$winners, freq = game_winners$n, min.freq = 0,
          max.words=200, random.order=FALSE, rot.per=0.3, 
          colors=brewer.pal(9, "GnBu")[3:9], scale=c(3,.5))
```


### 5.2 Regression models
#### 5.2.1 Games dataset
```{r}
reg <- lm(home_win ~ pts_home+pts_away+yds_home+yds_away+turnovers_home+turnovers_away, data = games)

summary(reg)
```

#### 5.2.2 Standings dataset (playoff as succes)
```{r}
reg2 <- lm(playoffs ~ wins+loss+points_for+points_against+points_differential+margin_of_victory+strength_of_schedule+simple_rating+offensive_ranking+defensive_ranking, data = standings)

summary(reg2)


```

#### 5.2.3 Standings dataset (Super Bowl win as success)
```{r}
reg3 <- lm(sb_winner ~ wins+loss+points_for+points_against+points_differential+margin_of_victory+strength_of_schedule+simple_rating+offensive_ranking+defensive_ranking, data = standings)

summary(reg3)


reg4 <- lm(sb_winner ~ wins+offensive_ranking+defensive_ranking, data = standings)

summary(reg4) #hypothesis 5


```


 
## 6 Prediction models
```{r}
RNGkind(sample.kind = "Rounding")
set.seed(356)


games$home_win <- as.factor(games$home_win)

# Training and test dataset  
games_index <- createDataPartition(games$home_win, p=0.8, list=FALSE) #Index

games_train <- games[games_index,]
games_train[ ,c('away_team_name', 'home_team_name', 'winners')] <- list(NULL)

games_test <- games[-games_index,]
games_test[ ,c('away_team_name', 'home_team_name', 'winners')] <- list(NULL)

```

### 6.1 Games dataset

#### 6.1.1 Naive Bayes
```{r}
model_nb <- train(games_train[, 3:8], games_train[, 2], method = 'naive_bayes')  #preProcess=c("center", "scale") 
predictions_nb <- predict(object = model_nb, games_test[,3:8])

# Confusion matrix 
confusionMatrix(predictions_nb, games_test[,2], positive = "1")

varimp_nb <- varImp(model_nb)
plot(varimp_nb, main="Variable Importance with naive bayes classifier")
```


#### 6.1.2 SVM (Support Vector Machine)
```{r}

model_svm<- train(games_train[, 3:8], games_train[, 2], method = 'svmLinear',preProcess=c("BoxCox")) #preProcess=c("center", "scale")
predictions_svm <- predict(object = model_svm, games_test[,3:8])

# Confusion matrix 
confusionMatrix(predictions_svm, games_test[,2], positive = "1")

varimp_svm <- varImp(model_svm)
plot(varimp_svm, main="Variable Importance with support vector machine classifier")


```


#### 6.1.3 K-nearest neighbour
```{r}
# Train a model using KNN algorithm
model_knn_norm <- train(games_train[, 3:8], games_train[, 2], method = 'knn',preProcess = c("range")) 
 
# Predict the labels of the test set
predictions_knn_norm <- predict(object = model_knn_norm, games_test[,3:8]) #the first 11 variables

# Confusion matrix 
confusionMatrix(predictions_knn_norm, games_test[,2], positive = "1") #the 12th variable is our outcome/predictive variable

varimp_knn_norm <- varImp(model_knn_norm)
plot(varimp_knn_norm, main="Variable Importance with k-neares neighbor classifier")

```


#### 6.1.4 Decision tree
```{r}

#Decision tree
model_dt <- train(games_train[, 3:8], games_train[, 2], method = 'rpart') #preProcess=c("center", "scale")

predictions_dt <- predict(object = model_dt, games_test[,3:8])

# Confusion matrix 
confusionMatrix(predictions_dt, games_test[,2], positive = "1")

#Variable importance plot
varimp_dt <- varImp(model_dt)
plot(varimp_dt, main="Variable Importance with decision tree classifier")

#Let's take a look at how the decision tree that the model has used to predict ended up looking!
prp(model_dt$finalModel, box.palette = "Reds", tweak = 1.2) #the "tweak" parameter controls text size, and "box.palette" controls node colour

```
















### 6.2 Standings dataset (playoffs as success parameter)
```{r}

standings$playoffs <- as.factor(standings$playoffs)
standings$sb_winner <- as.factor(standings$sb_winner)


```


```{r}

# Training and test dataset for standings with playoffs as success parameter

standings_index <- createDataPartition(standings$playoffs, p=0.8, list=FALSE) #Index

standings_train <- standings[standings_index,]

standings_test <- standings[-standings_index,]
```


#### 6.2.1 Naive Bayes
```{r}

model_nb2 <- train(standings_train[, 2:11], standings_train[, 12], method = 'naive_bayes')  #preProcess=c("center", "scale") 
predictions_nb2 <- predict(object = model_nb2, standings_test[,2:11])

# Confusion matrix 
confusionMatrix(predictions_nb2, standings_test[,12], positive = "1")

varimp_nb2 <- varImp(model_nb2)
plot(varimp_nb2, main="Variable Importance with naive bayes classifier")



str(standings)


HEY <- naive_bayes(playoffs ~ ., data = standings_train)
HEY

```



#### 6.2.2 SVM
```{r}
model_svm2 <- train(standings_train[, 2:11], standings_train[, 12], method = 'svmLinear',preProcess=c("BoxCox")) #preProcess=c("center", "scale")
predictions_svm2 <- predict(object = model_svm2, standings_test[,2:11])

# Confusion matrix 
confusionMatrix(predictions_svm2, standings_test[,12], positive = "1")

varimp_svm2 <- varImp(model_svm2)
plot(varimp_svm2, main="Variable Importance with support vector machine classifier")
```





#### 6.2.3 KNN
```{r}
# Train a model using KNN algorithm
model_knn_norm2 <- train(standings_train[, 2:11], standings_train[, 12], method = 'knn',preProcess = c("range")) 
 
# Predict the labels of the test set
predictions_knn_norm2 <- predict(object = model_knn_norm2, standings_test[,2:11]) #the first 11 variables

# Confusion matrix 
confusionMatrix(predictions_knn_norm2, standings_test[,12], positive = "1") #the 12th variable is our outcome/predictive variable

varimp_knn_norm2 <- varImp(model_knn_norm2)
plot(varimp_knn_norm2, main="Variable Importance with k-neares neighbor classifier")
```




#### 6.2.4 Decision tree
```{r}
#Decision tree
model_dt2 <- train(standings_train[, 2:11], standings_train[, 12], method = 'rpart') #preProcess=c("center", "scale")

predictions_dt2 <- predict(object = model_dt2, standings_test[,2:11])

# Confusion matrix 
confusionMatrix(predictions_dt2, standings_test[,12], positive = "1")

#Variable importance plot
varimp_dt2 <- varImp(model_dt2)
plot(varimp_dt2, main="Variable Importance with decision tree classifier")

#Let's take a look at how the decision tree that the model has used to predict ended up looking!
prp(model_dt2$finalModel, box.palette = "Reds", tweak = 1.2) #the "tweak" parameter controls text size, and "box.palette" controls node colour
```



















### 6.3 Standings dataset (winning superbowl as success parameter)
```{r}
# Training and test dataset for standings with playoffs as success parameter
standings_index2 <- createDataPartition(standings$sb_winner, p=0.8, list=FALSE) #Index

standings_train2 <- standings[standings_index2,]

standings_test2 <- standings[-standings_index2,]
```

#### 6.3.1 Naive Bayes
```{r}
model_nb3 <- train(standings_train2[, 2:11], standings_train2[, 13], method = 'naive_bayes')  #preProcess=c("center", "scale") 
predictions_nb3 <- predict(object = model_nb3, standings_test2[,2:11])

# Confusion matrix 
confusionMatrix(predictions_nb3, standings_test2[,13], positive = "1")

varimp_nb3 <- varImp(model_nb3)
plot(varimp_nb3, main="Variable Importance with naive bayes classifier")



str(standings)


HEY <- naive_bayes(playoffs ~ ., data = standings_train)
HEY
```


#### 6.3.2 SVM
```{r}
model_svm3 <- train(standings_train2[, 2:11], standings_train2[, 13], method = 'svmLinear',preProcess=c("BoxCox")) #preProcess=c("center", "scale")
predictions_svm3 <- predict(object = model_svm3, standings_test2[,2:11])

# Confusion matrix 
confusionMatrix(predictions_svm3, standings_test2[,13], positive = "1")

varimp_svm3 <- varImp(model_svm3)
plot(varimp_svm3, main="Variable Importance with support vector machine classifier")
```




#### 6.3.3 KNN
```{r}
# Train a model using KNN algorithm
model_knn_norm3 <- train(standings_train2[, 2:11], standings_train2[, 13], method = 'knn',preProcess = c("range")) 
 
# Predict the labels of the test set
predictions_knn_norm3 <- predict(object = model_knn_norm3, standings_test2[,2:11]) #the first 11 variables

# Confusion matrix 
confusionMatrix(predictions_knn_norm3, standings_test2[,13], positive = "1") #the 12th variable is our outcome/predictive variable

varimp_knn_norm3 <- varImp(model_knn_norm3)
plot(varimp_knn_norm3, main="Variable Importance with k-neares neighbor classifier")
```



#### 6.3.4 Decision tree
```{r}
#Decision tree
model_dt3 <- train(standings_train2[, 2:11], standings_train2[, 13], method = 'rpart') #preProcess=c("center", "scale")

predictions_dt3 <- predict(object = model_dt3, standings_test2[,2:11])

# Confusion matrix 
confusionMatrix(predictions_dt3, standings_test2[,13], positive = "1")

#Variable importance plot
varimp_dt3 <- varImp(model_dt3)
plot(varimp_dt3, main="Variable Importance with decision tree classifier")

#Let's take a look at how the decision tree that the model has used to predict ended up looking!
prp(model_dt3$finalModel, box.palette = "Reds", tweak = 1.2) #the "tweak" parameter controls text size, and "box.palette" controls node colour
```





